---
title: "R Ch 24 Section 3"
output: html_notebook
---



```{r setup, include=FALSE}
library(tidyverse)
library(modelr)
options(na.action = na.warn)

library(nycflights13)
library(lubridate)
```

#What affects the number of daily flights?

Let’s work through a similar process for a dataset that seems even simpler at first glance: the number of flights that leave NYC per day. This is a really small dataset — only 365 rows and 2 columns — and we’re not going to end up with a fully realised model, but as you’ll see, the steps along the way will help us better understand the data. Let’s get started by counting the number of flights per day and visualising it with ggplot2.

```{r}
daily <- flights %>% 
  mutate(date = make_date(year, month, day)) %>% 
  group_by(date) %>% 
  summarise(n = n())
daily
#> # A tibble: 365 × 2
#>         date     n
#>       <date> <int>
#> 1 2013-01-01   842
#> 2 2013-01-02   943
#> 3 2013-01-03   914
#> 4 2013-01-04   915
#> 5 2013-01-05   720
#> 6 2013-01-06   832
#> # ... with 359 more rows

ggplot(daily, aes(date, n)) + 
  geom_line()
```

##Day of the week

Understanding the long-term trend is challenging because there’s a very strong day-of-week effect that dominates the subtler patterns. Let’s start by looking at the distribution of flight numbers by day-of-week:

```{r}
daily <- daily %>% 
  mutate(wday = wday(date, label = TRUE))
ggplot(daily, aes(wday, n)) + 
  geom_boxplot()

```

There are fewer flights on weekends because most travel is for business. The effect is particularly pronounced on Saturday: you might sometimes leave on Sunday for a Monday morning meeting, but it’s very rare that you’d leave on Saturday as you’d much rather be at home with your family.

One way to remove this strong pattern is to use a model. First, we fit the model, and display its predictions overlaid on the original data:

```{r}
mod <- lm(n ~ wday, data = daily)

grid <- daily %>% 
  data_grid(wday) %>% 
  add_predictions(mod, "n")

ggplot(daily, aes(wday, n)) + 
  geom_boxplot() +
  geom_point(data = grid, colour = "red", size = 4)
```

Next we compute and visualise the residuals:

```{r}
daily <- daily %>% 
  add_residuals(mod)
daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

Note the change in the y-axis: now we are seeing the deviation from the expected number of flights, given the day of week. This plot is useful because now that we’ve removed much of the large day-of-week effect, we can see some of the subtler patterns that remain:

1. Our model seems to fail starting in June: you can still see a strong regular pattern that our model hasn’t captured. Drawing a plot with one line for each day of the week makes the cause easier to see:

```{r}
ggplot(daily, aes(date, resid, colour = wday)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

Our model fails to accurately predict the number of flights on Saturday: during summer there are more flights than we expect, and during Fall there are fewer. We’ll see how we can do better to capture this pattern in the next section.

2. There are some days with far fewer flights than expected:

```{r}
daily %>% 
  filter(resid < -100)
#> # A tibble: 11 × 4
#>         date     n  wday resid
#>       <date> <int> <ord> <dbl>
#> 1 2013-01-01   842  Tues  -109
#> 2 2013-01-20   786   Sun  -105
#> 3 2013-05-26   729   Sun  -162
#> 4 2013-07-04   737 Thurs  -229
#> 5 2013-07-05   822   Fri  -145
#> 6 2013-09-01   718   Sun  -173
#> # ... with 5 more rows
```

If you’re familiar with American public holidays, you might spot New Year’s day, July 4th, Thanksgiving and Christmas. There are some others that don’t seem to correspond to public holidays. You’ll work on those in one of the exercises.

3. There seems to be some smoother long term trend over the course of a year. We can highlight that trend with geom_smooth():

```{r}
daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line(colour = "grey50") + 
  geom_smooth(se = FALSE, span = 0.20)
#> `geom_smooth()` using method = 'loess'
```

There are fewer flights in January (and December), and more in summer (May-Sep). We can’t do much with this pattern quantitatively, because we only have a single year of data. But we can use our domain knowledge to brainstorm potential explanations.

##Seasonal saturday effect

Let’s first tackle our failure to accurately predict the number of flights on Saturday. A good place to start is to go back to the raw numbers, focussing on Saturdays:

```{r}
daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n)) + 
    geom_point() + 
    geom_line() +
    scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

(I’ve used both points and lines to make it more clear what is data and what is interpolation.)

I suspect this pattern is caused by summer holidays: many people go on holiday in the summer, and people don’t mind travelling on Saturdays for vacation. Looking at this plot, we might guess that summer holidays are from early June to late August. That seems to line up fairly well with the state’s school terms: summer break in 2013 was Jun 26–Sep 9.

Why are there more Saturday flights in the Spring than the Fall? I asked some American friends and they suggested that it’s less common to plan family vacations during the Fall because of the big Thanksgiving and Christmas holidays. We don’t have the data to know for sure, but it seems like a plausible working hypothesis.

Lets create a “term” variable that roughly captures the three school terms, and check our work with a plot:

```{r}
term <- function(date) {
  cut(date, 
    breaks = ymd(20130101, 20130605, 20130825, 20140101),
    labels = c("spring", "summer", "fall") 
  )
}

daily <- daily %>% 
  mutate(term = term(date)) 

daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n, colour = term)) +
  geom_point(alpha = 1/3) + 
  geom_line() +
  scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

(I manually tweaked the dates to get nice breaks in the plot. Using a visualisation to help you understand what your function is doing is a really powerful and general technique.)

It’s useful to see how this new variable affects the other days of the week:

```{r}
daily %>% 
  ggplot(aes(wday, n, colour = term)) +
    geom_boxplot()
```

It looks like there is significant variation across the terms, so fitting a separate day of week effect for each term is reasonable. This improves our model, but not as much as we might hope:

```{r}
mod1 <- lm(n ~ wday, data = daily)
mod2 <- lm(n ~ wday * term, data = daily)

daily %>% 
  gather_residuals(without_term = mod1, with_term = mod2) %>% 
  ggplot(aes(date, resid, colour = model)) +
    geom_line(alpha = 0.75)
```

We can see the problem by overlaying the predictions from the model on to the raw data:

```{r}
grid <- daily %>% 
  data_grid(wday, term) %>% 
  add_predictions(mod2, "n")

ggplot(daily, aes(wday, n)) +
  geom_boxplot() + 
  geom_point(data = grid, colour = "red") + 
  facet_wrap(~ term)
```

Our model is finding the mean effect, but we have a lot of big outliers, so mean tends to be far away from the typical value. We can alleviate this problem by using a model that is robust to the effect of outliers: MASS::rlm(). This greatly reduces the impact of the outliers on our estimates, and gives a model that does a good job of removing the day of week pattern:

```{r}
mod3 <- MASS::rlm(n ~ wday * term, data = daily)

daily %>% 
  add_residuals(mod3, "resid") %>% 
  ggplot(aes(date, resid)) + 
  geom_hline(yintercept = 0, size = 2, colour = "white") + 
  geom_line()
```

It’s now much easier to see the long-term trend, and the positive and negative outliers.

##Computed variables

If you’re experimenting with many models and many visualisations, it’s a good idea to bundle the creation of variables up into a function so there’s no chance of accidentally applying a different transformation in different places. For example, we could write:

```{r}
compute_vars <- function(data) {
  data %>% 
    mutate(
      term = term(date), 
      wday = wday(date, label = TRUE)
    )
}
```

Another option is to put the transformations directly in the model formula:

```{r}
wday2 <- function(x) wday(x, label = TRUE)
mod3 <- lm(n ~ wday2(date) * term(date), data = daily)
```

Either approach is reasonable. Making the transformed variable explicit is useful if you want to check your work, or use them in a visualisation. But you can’t easily use transformations (like splines) that return multiple columns. Including the transformations in the model function makes life a little easier when you’re working with many different datasets because the model is self contained.

##Time of the year: an alternative approach

In the previous section we used our domain knowledge (how the US school term affects travel) to improve the model. An alternative to using our knowledge explicitly in the model is to give the data more room to speak. We could use a more flexible model and allow that to capture the pattern we’re interested in. A simple linear trend isn’t adequate, so we could try using a natural spline to fit a smooth curve across the year:

```{r}
library(splines)
mod <- MASS::rlm(n ~ wday * ns(date, 5), data = daily)

daily %>% 
  data_grid(wday, date = seq_range(date, n = 13)) %>% 
  add_predictions(mod) %>% 
  ggplot(aes(date, pred, colour = wday)) + 
    geom_line() +
    geom_point()
```

We see a strong pattern in the numbers of Saturday flights. This is reassuring, because we also saw that pattern in the raw data. It’s a good sign when you get the same signal from different approaches.

##Questions
###**Question 4**
*Create a new variable that splits the wday variable into terms, but only for Saturdays, i.e. it should have Thurs, Fri, but Sat-summer, Sat-spring, Sat-fall. How does this model compare with the model with every combination of wday and term?*

```{r}
holidays <- c(20130101, # new years
             20130121, # mlk
             20130218, # presidents
             20130527, # memorial
             20130704, # independence
             20130902, # labor
             20131028, # columbus
             20131111, # veterans
             20131128, # thanksgiving
             20131225)
holidays1 <- holidays %>% 
  lubridate::ymd()

daily2 <- daily %>% 
  mutate(term_holiday = case_when(
    .$date %in% holidays1 ~"holiday",
    .$wday == "Sat" & .$term == "summer" ~ "Sat-summer",
    .$wday == "Sat" & .$ term == "fall" ~ "Sat-fall",
    .$wday == "Sat" & .$term == "spring" ~ "Sat-spring",
    TRUE ~ as.character(.$wday)))

mod_term_hol <- lm(n ~ term_holiday, data = daily2)

daily_res <- daily2 %>% 
  add_residuals(mod_term_hol, "resid") %>% 
  ggplot(aes(date, resid))+
  geom_line()

daily_res

```

###**Question 8**
*It’s a little frustrating that Sunday and Saturday are on separate ends of the plot. Write a small function to set the levels of the factor so that the week starts on Monday.*

```{r}
Sunday_last <- function(x) {
  fct_relevel(x, "Sun", after = 6)
}
```









